---
title: "TEOS Hydrostation S Data Lab"
author: "Jenniffer Perez"
date: "`r Sys.Date()`"
output: 
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(gsw)
```

Here is **The Gibbs Sea Water (GSW) Oceanographic data of Thermodynamic Equation of Seawater (TEOS-10)**. This data can be found in: <http://www.teos-10.org/pubs/gsw/html/gsw_contents.html>. For this assignment we observed the data for dissolved oxygen over seasons for the Bermuda. Winter months (December to March) and Summer months (August to October) were observed. The object of this assignment was to observe if there were changes in the dissolved oxygen through winter and summer seasons. It was hypothesized that in the Bermuda, there will be more high levels of dissolved oxygen in the winter than in the summer.  A simple linear model was applied to create a regression model for the data selected for the Bermuda assignment.   The results for the p-value was less than 2.2e-16, this results indicate that there is no significant difference in dissolved oxygen for the Winter and Summer season in the Bermuda for the first 100m of water depth.


```{r, echo=FALSE, warning=FALSE, message=FALSE}
hydrostation_bottle <- read_delim("hydrostation_bottle.txt", 
    delim = "\t", escape_double = FALSE, 
    col_names = FALSE, trim_ws = TRUE, skip = 31)
View(hydrostation_bottle)


hydrostation_bottle_names <- read_csv("hydrostation_bottle.txt", 
    skip = 30)
View(hydrostation_bottle_names)

colnames(hydrostation_bottle)=colnames(hydrostation_bottle_names)

#view(hydrostation_bottle)

```

## Variables Names and Units for the TEOS-10 data.

-   yyyymmdd = Year Month Day\
-   decy = Decimal Year\
-   time = Time (hhmm)\
-   latN = Latitude (Deg N)
-   lonW = Longitude (Deg W)
-   Depth = Depth (m)\
-   Temp = Temperature ITS-90 (C)
-   Pres = CTD Pressure (dbar)\
-   CTD_S = CTD Salinity (PSS-78)\
-   Sal1 = Salinity-1 (PSS-78)\
-   Sig-th = Sigma-Theta (kg/m\^3)\
-   O2(1) = Oxygen-1 (umol/kg)\
-   OxFixT = Oxygen Fix Temp (C)\
-   Anom1 = Oxy Anomaly-1 (umol/kg)\
-   /Quality flags
-   -999 = No data
-   0 = Less than detection limit

```{r}
# Plotting the data.  

hydrostation_bottle %>% 
 ggplot()+geom_point(aes(x=decy,y=`Sig-th`)) 
  

#Plotting the data for a better intrepretation. 

hydrostation_bottle %>%
  filter(`Sig-th`!=-999) %>%
  ggplot()+ geom_point(aes(x=decy,y=`Sig-th`))

#Plotting the data for a linear plot.

hydrostation_bottle %>%
  filter(`Sig-th` !=-999 & Depth <20) %>%
  ggplot()+geom_line(aes(x=decy,y=`Sig-th`))

#Clear seasonal signal for sigma theta. lets see how this compares to temperature.

hydrostation_bottle %>%
  filter(`Sig-th` !=-999 & Depth <20) %>%
  ggplot()+geom_point(aes(x=Temp,y=`Sig-th`))

#Temperature and density are strongly correlates, but there appears to be 2 outlines that we will likely need to address at some point.

#We only have density data from 1988-present, but temp and salinity data from 1950s- present. 
#This means I can calculate seawater density from 1950s to present.  
#TEOS-10 will be used to do this..

```

## Teos-10 Toolbox in Package seacarb

```{r}
?gsw 

#Launches the documentation for the gibbs seawater toolbox (TEOS-10)

?gsw_sigma0
#It says we need absolute salinity and conservative temperature

#First we need absolute salinity.

?gsw_SA_from_SP
#practical salinity
#sea pressure (dbar)
#longitude 
#latitude

#Let's plot our pressure data - it's missing before 1980's

hydrostation_bottle %>%
  ggplot()+geom_point(aes(x=decy,y=Pres))

#We have depth data for the time series
hydrostation_bottle %>%
  ggplot()+
  geom_point(aes(x=decy,y=Depth))

?gsw_p_from_z

#adds a pressure column from/to the depth and latN columns from hydrostation bottle

hydrostation_bottle= hydrostation_bottle %>%
  mutate(Pres_gsw=gsw_p_from_z(Depth*-1,latN))

#mutate() creates a new column that are functions of exisisting variables.It can also modify (if the name is the same as an existing column) and delete columns (by setting their value to NULL ).


hydrostation_bottle %>%
  ggplot()+
  geom_point(aes(x=Pres,y=Pres_gsw))

#checking lat long and salinity data

hydrostation_bottle %>%
  ggplot()+
  geom_point(aes(x=decy,y=latN))

hydrostation_bottle %>%
  ggplot()+
  geom_point(aes(x=decy,y=lonW))

hydrostation_bottle %>%
  ggplot()+
  geom_point(aes(x=decy,y=CTD_S))

hydrostation_bottle %>%
  ggplot()+
  geom_point(aes(x=decy,y=Sal1))

hydrostation_bottle= hydrostation_bottle %>%
  mutate(Pres_gsw=gsw_p_from_z(Depth*-1,latN)) %>%
  mutate(S_abs_gsw=gsw_SA_from_SP(Sal1,Pres_gsw,360-lonW,latN))

#check it
hydrostation_bottle %>%
  ggplot()+
  geom_point(aes(x=decy,y=S_abs_gsw))
#how else can I check my data?

hydrostation_bottle %>%
  filter(Sal1!=-999) %>%
  ggplot()+
  geom_point(aes(x=Sal1,y=S_abs_gsw))

#now we need to calculate the conservative temp

?gsw_CT_from_t

#We need absolute salinity, in-situ temp (ITS-90), and sea pressure

#add line to calculate conservative temp
HydroS= hydrostation_bottle %>%
 filter(Sal1!=-999) %>%
  mutate(Pres_gsw=gsw_p_from_z(Depth*-1,latN)) %>%
  mutate(S_abs_gsw=gsw_SA_from_SP(Sal1,Pres_gsw,360-lonW,latN)) %>%
  mutate(T_cons_gsw=gsw_CT_from_t(S_abs_gsw,Temp,Pres_gsw))

#let's check our data

HydroS %>%
  filter(Temp!=-999) %>%
  ggplot()+
  geom_point(aes(x=Temp,y=T_cons_gsw))

#add line to calculate conservative temperature

HydroS = hydrostation_bottle %>%  #replace the homeworkline here select? may be usefull or combination of filters.
  filter(Sal1!=-999) %>%
  filter(Temp!=-999) %>%
  mutate(Pres_gsw=gsw_p_from_z(Depth*-1,latN)) %>%
  mutate(S_abs_gsw=gsw_SA_from_SP(Sal1,Pres_gsw,360-lonW,latN)) %>%
  mutate(T_cons_gsw=gsw_CT_from_t(S_abs_gsw,Temp,Pres_gsw))
  
#add line to calculate conservative temperature

  HydroS = hydrostation_bottle %>%
  filter(Sal1!=-999) %>%
  filter(Temp!=-999) %>%
  mutate(Pres_gsw=gsw_p_from_z(Depth*-1,latN)) %>%
  mutate(S_abs_gsw=gsw_SA_from_SP(Sal1,Pres_gsw,360-lonW,latN)) %>%
  mutate(T_cons_gsw=gsw_CT_from_t(S_abs_gsw,Temp,Pres_gsw)) %>%
  mutate(Sig_th_gsw=gsw_sigma0(S_abs_gsw,T_cons_gsw))
  
  #our bottle and ctd salinity do not agree, this is likely the problem for the los-sig-th-gsw

  HydroS%>%
  filter(Sig_th_gsw<0) %>% 
    view()
  
  
  HydroS %>%
  filter(Sig_th_gsw<0) %>% 
  mutate(S_abs_gsw=gsw_SA_from_SP(CTD_S,Pres_gsw,360-lonW,latN)) %>%
  mutate(T_cons_gsw=gsw_CT_from_t(S_abs_gsw,Temp,Pres_gsw)) %>%
  mutate(Sig_th_gsw=gsw_sigma0(S_abs_gsw,T_cons_gsw)) 
  #view()
  
 HydroS_correctedS_a=HydroS %>%
  filter(Sig_th_gsw<0) %>% 
   mutate(S_abs_gsw=gsw_SA_from_SP(CTD_S,Pres_gsw,360-lonW,latN)) %>%
  mutate(T_cons_gsw=gsw_CT_from_t(S_abs_gsw,Temp,Pres_gsw)) %>%
  mutate(Sig_th_gsw=gsw_sigma0(S_abs_gsw,T_cons_gsw)) 

 HydroS_correctedS_b=HydroS %>%
    filter(Sig_th_gsw>0) 
 
 HydroS_corrected=rbind(HydroS_correctedS_a,HydroS_correctedS_b)
 
#row bind function represents a row bind function for vectors, data frame, and matrices to be arranged as rows.  It uses combine multiple data frames for data manipulation
 
 HydroS_corrected %>% 
   filter(`Sig-th`!=-999) %>% 
   ggplot()+geom_point(aes(x=`Sig-th`,y=Sig_th_gsw))
 
 HydroS_corrected %>% 
   ggplot()+geom_point(aes(x=Sig_th_gsw,y=Depth))+scale_y_reverse()+scale_x_continuous(position="top")+xlab("Depth (m)")+theme_classic()
 

```

## Has surface sigma theta decreased over time?

```{r}
HydroS_shallow = HydroS_corrected %>% 
  filter(Depth<30)

?lm

lm(Sig_th_gsw~decy,data = HydroS_shallow)

#coefficients (intercept and decy)

#y=mx+b

#y = sig_th_gsw

#x = decy

#coefficients: intercept = b, decy = m

#Sig_th_gsw = -0.004*decy + 33.4

#(kg/m^3/y)*y + (kg/m^3)

sig_theta_time_model=lm(Sig_th_gsw~decy, data=HydroS_shallow) 

summary(sig_theta_time_model)

library(plotly)

plot=HydroS_shallow %>% 
  ggplot(aes(x=decy,y=Sig_th_gsw))+geom_point()+geom_line()+geom_smooth(method="lm")+ theme_classic()

ggplotly(plot)



```

##Asignment in class (March 16)

```{r}

#In Bermuda
#Dec - March = Winter Months 
#August to October = Summer Months

HydroS_seasons=HydroS_corrected %>% 
  mutate(month=as.numeric(substr(yyyymmd,5,6))) %>% 
  mutate(season=ifelse(month==12|month==1|month==2|month==3,'winter',ifelse(month==8|month==9|month==10,'summer',NA)))
  
#Is dissolved oxygen higher in summer than winter?
#Is sainity higher in summer than wint#is sound speed higher....
#is spiciness higher....

summary(lm(`O2(1)`~season,data=HydroS_seasons))


```

```{r}
HydroS_seasons=HydroS_corrected %>% 
  filter(Depth<100) %>% 
  filter(`O2(1)`!=-999) %>% 
  mutate(month=as.numeric(substr(yyyymmd,5,6))) %>% 
  mutate(season=ifelse(month==12|month==1|month==2|month==3,'winter',ifelse(month==8|month==9|month==10,'summer',NA))) %>% 
  filter(season=="winter"|season=="summer")
    
  
  
#Is dissolved oxygen higher in winter than summer? 

#filter for 100m
 #jitterplot points spread apart add a box plot
  #there is no difference in the surface meters

summary(lm(`O2(1)`~season,data=HydroS_seasons))

HydroS_seasons %>% 
  ggplot(aes(x=season, y=`O2(1)`))+geom_point()+geom_jitter()+geom_boxplot()+labs(caption="Plot for dissolved oxygen per Winter and Summer seasons for 100m of depth for the Bermuda.")


```
